### 2018-09-29 paper/paper.tex
Detailed revisions

% Please keep the abstract below 300 words
\section*{Abstract}
When it comes to folding, the RNA energy landscape is particularly rugged, densely populated with 
energy-favorable pairings of nucleotide sequences, most of which are not a part of the molecule's native structure or structures. Yet many non-coding RNA molecules spontaneously fold to a native structure when placed in a suitable chemical environment at an appropriate temperature.
Levinthal's convincing arguments (aka Levinthal's paradox) about proteins apply to RNA molecules as well, perhaps even more compellingly: arrival to thermal equilibrium should take a great deal more time than the biological lifetime of the organism. Evidently, mechanisms have evolved that contribute to a directed and efficient folding process, and quite possibly the active configuration is not at all like a sample from an equilibrium distribution, but rather a metastable state, provided only that it will typically last long enough for its purpose. 
Various measures of the difficulty in reaching a stable configuration, whether or not it is a true equilibrium,
are based on the number of possible Watson-Crick or G$\cdot$U wobble pairings at each location. We call these measures ``ambiguity indexes,'' and focus on two, each of which is a function of both the primary and any given proposed secondary structure. Comparing families of RNA molecules that appear in ribonucleoproteins (``bound'' molecules) to RNA molecules that do not operate as nucleoproteins (``unbound'' molecules), we find that the unbound molecules have systematically lower ambiguity indexes, {\em provided that the indexes are computed using the consensus structure} (also known as the ``gold standard''). The effect largely disappears when the indexes are based, instead, on the so-called minimum-free energy (MFE) structures. Since MFE structures are derived from an approximation of the equilibrium distribution, we interpret these and other related results as evidence for metastable but not necessarily thermal equilibrium secondary structure. 
%
%
%% Please keep the Author Summary between 150 and 200 words
%% Use first person. PLOS ONE authors please skip this step. 
%% Author Summary not valid for PLOS ONE submissions.   
%\section*{Author Summary}
%Recent discoveries indicate that, in addition to being a messenger between DNA and protein, RNA molecules assume a wide range of biological functions. For biological macromolecules, structure is function. Experimental determination of RNA structures is still time-consuming, and computational approaches are of great importance. The prevailing computational approach tries to find the structure with the minimum energy, yet the relevance of this minimum energy structure to the native structure is still hotly debated. In this paper, we adopt a kinetic perspective, and argue that more emphasis should be placed on the folding process when trying to develop computational methods for RNA structure prediction. We present some statistical analyses using the primary and secondary data (sequence and base-pairs data) of RNA molecules, based on the concept of ``local ambiguity", i.e. the molecule's tendency to ``make a mistake" at a certain location when forming secondary structures. Our results highlight deficiencies in the minimum-energy approach, and demonstrate the importance of considering kinetics as well as protein-RNA interactions in developing computational approaches for RNA secondary structure prediction.

@@

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}

@@

Because of the time-consuming nature of experimental determination of RNA structures, a considerable amount of work has been put into computational (as opposed to experimental) approaches. For secondary structure prediction, comparative analysis is the gold standard  \cite{Gutell1992-hu}. But accurate comparative analysis typically requires a large number of homologous sequences, which may or may not be available. On the other hand, the prevailing {\em computational} approaches  xxx References???xxx are based on an important and controversial assumption. Namely, that the structures of  non-coding RNA molecules, {\em in vivo}, are in thermal equilibrium. If this were the case, and if the configuration energies could be accurately specified, then secondary (and in principle, even tertiary) structures could be explored through Monte Carlo sampling from the Boltzmann equilibrium distribution. Alternatively, depending on the nature of the energy landscape, secondary structures could be approximated by identifying the minimum free energy (MFE) configuration  \cite{Mathews1999-hc,Zuker1999-rc}. However, even if we put aside the necessity for approximations, the biological relevance of equilibrium configurations has been a source of misgivings at least since 1969, when Levinthal pointed out that the time required to equilibrate might be too long by many orders of magnitude  \cite{Levinthal1969-hi}. In light of these observations, and considering the ``frustrated'' nature of the folding landscape, many have argued that when it comes to structure prediction for macromolecules, kinetic accessibility is more relevant than equilibrium thermodynamics  \cite{Higgs2000-gj,Flamm2008-ir,Baker1994-px}.

@@

We will introduce quantitative measures of ambiguity and demonstrate their statistical relationships to native secondary structure and to qualitative distinctions between RNA families.
Using a somewhat arbitrary convention, we will
refer to sequences of length four (four consecutive nucleotides) as {\em segments}. Keeping this convention in mind, each line of statistical
evidence is based on the same formal definition of the ``local ambiguity'' of a given segment. This is simply the number of its complementary pairs in the molecule. 
When we refer to the {\em location} of a segment, we will mean the location of the first element of the segment, counting from 5$^\prime$ to 3$^\prime$, and  
when we refer to the local ambiguity of a location, we will mean the local ambiguity of the segment at that location. 
Local ambiguity is an intrinsic property of the primary structure. We will be interested in exposing its statistical relationships to any given candidate secondary structure of the same molecule. For this purpose, for any particular secondary structure we distinguish three different kinds of locations:
\begin{description}
	\item[\textit{Single}:] Locations where all nucleotides in the corresponding segment are unpaired in the secondary structure;
	\item[\textit{Double}:] Locations where all nucleotides in the corresponding segment are paired in the secondary structure;
	\item[\textit{Transitional}:] Locations where some nucleotides in the corresponding segment are paired and others are unpaired in the secondary structure.
\end{description}
{\em Double} and {\em transitional} locations participate in the candidate secondary structure, while {\em single} locations do not.
As a general rule, local ambiguities measured at any of these locations will increase with the length of the molecule---there are simply more opportunities to find complementary sequences.
Instead of using the local ambiguities themselves, we focus on the {\em differences} between ambiguities in and around stems ({\em double} and {\em transitional} locations) from those at unpaired ({\em single}) locations. In particular, we defined the ``T-S ambiguity index'' to be the difference between the average local ambiguity at {\em transitional} locations minus the average at {\em single} locations. We defined the ``D-S ambiguity index'' similarly, but adjusted for an evident bias at {\em double} locations, most of which will have at least one complementary pair somewhere in the molecule. 
(The exceptions being bulges, which occur in about 20\% of the stems in our data set.)
With this in mind, we defined the D-S ambiguity index to be the average local ambiguity at {\em double} locations minus the average over those {\em single} locations that have at least one ambiguity\footnote{
An argument could be made for including bulges in the definition of the {\em local ambiguity} in the first place. We have not explored this or other variations, though there is not shortage of arguably useful descriptive statistics for capturing the notion of ambiguity.}.

@@

\subsection*{A First Look at the Data: Shuffling Nucleotides}
Our goals are to explore connections between ambiguities and basic characteristics of RNA families, as well as the changes in these relationships, if any, when using comparative, as opposed to MFE, secondary structures.
For each molecule and each location $i$, the segment at $i$ has been assigned a ``local ambiguity'' $a_i(p)$ that depends only on the primary structure, and a label ({\it single}, {\it double}, or {\it transitional}) that depends only on the secondary structure. Since the local ambiguity, by itself, is strongly dependent on the length of the molecule, and possibly on other intrinsic properties,
we propose two {\em relative} ambiguity indexes: T-S and D-S, each of which depends on both the primary ($p$) and purported secondary ($s$) structures. Formally,
\begin{equation}
	d_\text{T-S}(p, s) = \frac
	{\sum_{j = 0}^{N - 3} a_j (p) 
	c^{\text{tran}}_j (s)}
	{\sum_{j = 0}^{N - 3} 
	c^{\text{tran}}_j (s)} - 
	\frac
	{\sum_{j = 0}^{N - 3} a_j (p) c_j^{\text{single}} (s)}
	{\sum_{j = 0}^{N - 3} c_j^{\text{single}} (s)}
\end{equation}
where we have used
$c_i^{\text{tran}}$ and $c_i^{\text{single}}$ for indicating whether location $i$ is {\it transitional} or {\it single} respectively. In other words, for each $i=1,2,\ldots,N-3$
\begin{align}
	c_i^{\text{tran}} &=
	\begin{cases}
		1, & \text{if location $i$ is \textit{transitional}} \\
		0, & \text{otherwise}
	\end{cases} \\
	c_i^{\text{single}} &=
	\begin{cases}
		1, & \text{if location $i$ is \textit{single}} \\
		0, & \text{otherwise}
	\end{cases}
\end{align}
In short, the T-S ambiguity index is the difference in the averages of the local ambiguities at {\em transitional} sites and {\em single} sites.
Since, as noted earlier, the local ambiguity at every {\em double} location, $i$, 
is likely to be at least one ($a_i\geq 1$), the D-S index involves only those {\em single} locations, $j$, for which $a_j\geq 1$:
\begin{equation}
	d_\text{D-S}(p, s) = \frac
	{\sum_{j = 0}^{N - 3} a_j (p) 
	c^{\text{double}}_j (s)}
	{\sum_{j = 0}^{N - 3} 
	c^{\text{double}}_j (s)} - 
	\frac
	{\sum_{j = 0}^{N - 3} a_j (p) \hat{c}_j^{\text{single}} (p,s)}
	{\sum_{j = 0}^{N - 3} \hat{c}_j^{\text{single}} (p,s)}
\end{equation}
Here, $c_i^{\text{double}}=1$ if $i$ is {\em double}, and zero otherwise, and
$\hat{c}^{\text{single}}_i$ indicates those {\em single} locations with at least one pair: for each $i=1,2,\ldots,N-3$,
\begin{align}
	\hat{c}_i^{\text{single}} & =
	\begin{cases}
		1, & \text{if location $i$ is \textit{single} and $a_i(p)\geq 1$} \\
		0, & \text{otherwise}
	\end{cases}
\end{align}
Note that $\hat{c}^{\text{single}}_i$ depends on both primary and secondary structures: 
$\hat{c}^{\text{single}}_i$ $=$ $\hat{c}^{\text{single}}_i(p,s)$.

@@

To make this precise, consider an RNA molecule with primary structure $p$ and comparative secondary structure $s$. Construct a segment ``histogram function,'' $\mathcal{H}(p)$, which is the 
vector of the number of times that each of the $4^4$ possible segments appears in $p$.\footnote{${\cal H}$ is made a vector by fixing an arbitrary ordering of all possible segments.}.
Let $\mathcal{P}(p)$ be the set of all permutations of the ordering of nucleotides in $p$, and let 
$\mathcal{E}(p)\subseteq\mathcal{P}(p)$ be the subset of permutations that preserve the frequencies of four-tuples:
\[
{\cal E}(p) = \{p'\in\mathcal{P}(p): {\cal H}(p')={\cal H}(p)\}
\]
Clever algorithms exist for efficiently drawing independent samples from the uniform distribution on $\mathcal{E}$---see \cite{Kandel1996-kz, Fitch1983-ec, Altschul1985-nq}. Let $p^{(1)},
\ldots,p^{(K)}$ be $K$ such samples, and let $\displaystyle
d_\text{T-S}\big(p^{(1)},s\big),\ldots,d_\text{T-S}\big(p^{(K)},s\big)$ and
$\displaystyle
d_\text{D-S}\big(p^{(1)},s\big),\ldots,d_\text{D-S}\big(p^{(K)},s\big)$ 
be the corresponding T-S and D-S ambiguity indexes. The results are not vary sensitive to $K$, nor to the particular sample, provided that $K$ is large enough. We used $K=$10,000. Finally, let 
$\alpha_\text{T-S}(p,s) \in [0,1]$ and
$\alpha_\text{D-S}(p,s) \in [0,1]$
be the left-tail empirical probabilities, under the distributions defined by the ensembles
\begin{align}
&  d_\text{T-S}(p,s),d_\text{T-S}\big(p^{(1)},s\big),\ldots,d_\text{T-S}\big(p^{(K)},s\big) 
\label{eqn:T-Sensemble}\\
\text{and \hglue .8cm} 
& d_\text{D-S}(p,s),d_\text{D-S}\big(p^{(1)},s\big),\ldots,d_\text{D-S}\big(p^{(K)},s\big) 
\label{eqn:D-Sensemble}
\end{align}
of choosing an ambiguity index less than or equal to $d_\text{T-S}(p,s)$ and 
$d_\text{T-S}(p,s)$, respectively:
\begin{align}
\alpha_\text{T-S}(p,s)  & = \frac
{1 + \#\{k\in\{1,\ldots,K\}: d_\text{T-S}\big(p^{(k)},s\big) \leq d_\text{T-S}(p,s)\}}{1+K} \label{eqn:alpha_T-S} \\
\alpha_\text{D-S}(p,s)  & = \frac
{1 + \#\{k\in\{1,\ldots,K\}: d_\text{D-S}\big(p^{(k)},s\big)\leq d_\text{D-S}(p,s)\}}{1+K} \label{eqn:alpha_D-S}
\end{align}
In essence, each $\alpha$ score is a self-calibrated ambiguity index.

@@

We are given a population of $M$ molecules, $m=1,\ldots,M$, each with a binary outcome measure $B_m\in\{-1,+1\}$.
There are two subpopulations of interest: the first $M_1$ molecules make up population 1 and the next $M_2$ molecules make up population 2; $M_1+M_2=M$. We observe $n_1$ plus values in population 1 and $n_2$ in population 2 
\begin{align}
n_1 & = \#\big\{m\in\{1,2,\ldots,M_1\}:B_m=+1\big\} \\
n_2 & = \#\big\{m\in\{M_1+1,M_1+2,\ldots,M\}:B_m=+1\big\}
\end{align}
We suspect that population 1 has less than its share of plus ones, meaning that the $n_1+n_2$ population of plus ones was not randomly distributed among the $M$ molecules. To be precise, let $N$ be the number of plus ones that appear from a draw, without replacement, of $M_1$ samples from $B_1,\ldots,B_M$. Under the null hypothesis, $H_o$, $n_1$ is a sample from the hypergeometric distribution on $N$:
\begin{equation}
\label{eqn:null}
\mathbb{P}\{N=n\} = \frac{\binom{M_1}{n}\binom{M_2}{n+1+n_2-n}}
{\binom{M}{n_1+n_2}}\quad \max\{0,n_1+n_2-M_2\}\leq n 
\leq \min\{n_1+n_2,M_1\}
\end{equation}
The alternative hypothesis, $H_a$, is that $n_1$ is too small to be consistent with $H_o$, leading to a left-tail test with p-value $\mathbb{P}\{N\leq n_1\}$ (which can be computed directly or using a statistical package, e.g. {\em hygecdf} in MatLab).

@@

The comparative secondary structure of an RNA molecule, when combined with its primary structure, can be used to construct a measure---the T-S ambiguity index---which distinguishes unbound from bound molecules with good accuracy. Can the same can be said for the D-S index?
Yes, albeit with lower accuracy. To demonstrate, we followed the identical procedure, except that we assigned the index $d_\text{D-S}(p,s)$ rather than $d_\text{T-S}(p,s)$ to each molecule.  The ROC curve (with area 0.65) is shown in the lower-left panel of Figure \ref{fig:UnboundVSBound}. The hypergeometric test, based on the signs of the D-S indexes
(positive counts, for each RNA family, can be found in Table \ref{table:NumberPositive}) has p-value $7.3\cdot 10^{-8}$. Finally, the right-hand panels in Figure \ref{fig:UnboundVSBound} mirror the left-hand panels, except that all secondary structures were computed by (approximately) maximizing free energy rather than comparative analysis. The classification results are substantially less convincing and the p-values substantially higher.

@@

Interestingly, there is a sharp difference in predicting the source of secondary structure in unbound molecules as opposed to bound molecules. Consider the top two ROC curves in Figure
\ref{fig:CompVSMFE}. In each of the two experiments a classifier was constructed by thresholding the T-S ambiguity index, declaring the secondary structure to be 
`comparative' when $d_\text{T-S}(p,s)<t$ and `MFE' otherwise, since, as we have seen, smaller values of the ambiguity indexes are generally associated with the comparative secondary structure. The ROC curves are then swept out by varying $t$. The difference between the two panels is in the population used for the classification experiments---unbound molecules in the upper-left and bound molecules in the upper-right. In unbound molecules, small values of the T-S ambiguity index are a good indication that a secondary structure was derived from comparative rather than free-energy analysis, but not in bound molecules. The corresponding hypothesis tests seek evidence against the null hypotheses that in a given group (unbound or bound) the set of positive T-S ambiguity indexes ($d_\text{T-S}(p,s)>0$) are equally distributed between the comparative and free-energy derived indexes, and in favor of the alternatives that the T-S ambiguity indexes are less typically positive for the comparative secondary structures. The necessary data can be found in Table \ref{table:NumberPositive}. The results are consistent with the classification experiments: the hypergeometric p-value is 
is $5.4\cdot 10^{-14}$ for the unbound population and $0.07$ for the bound population.

@@

\section*{Discussion}
We have introduced two indexes designed as coarse measures of the prevalence of kinetic traps in the energy landscape of a structural RNA molecule. These indexes, which we call the T-S and D-S ambiguity indexes, are each a function of both the primary and any presumed secondary structure of the molecule. The first, the T-S index, compares the availability of false pairings (i.e. stems that are not part of the purported secondary structure) of sequences straddling transitions from double to single-stranded regions to the availability of false pairings strictly within single-stranded regions.
The second, the D-S index, results from a similar calculation, but using double-stranded sequences in place of {\em transitional} sequences. We have given evidence that either index can be used effectively to classify unbound RNA (molecules that do not act as components of ribonucleoproteins) from bound RNA (those that are components of ribonucleoproteins), {\em provided that the secondary structure is derived from comparative analysis}. 
Specifically, unbound molecules typically have lower ambiguity indexes than bound molecules. On the other hand, there is little evidence for this effect in the ambiguity indexes derived instead from MFE analyses, and classifiers based on MFE secondary structure are substantially less accurate. Furthermore, either index can be used in unbound molecules to effectively classify a given secondary structure as derived from comparative as opposed to MFE analysis, but not in bound molecules. In particular, among unbound molecules comparative analysis structures typically have lower ambiguity indexes than MFE structures.

>>>>

% Please keep the abstract below 300 words
\section*{Abstract}
When it comes to folding, the RNA energy landscape is particularly rugged, densely populated with 
energy-favorable pairings of nucleotide sequences, most of which are not a part of the molecule's native structure or structures. Yet many non-coding RNA molecules spontaneously fold to a native structure when placed in a suitable chemical environment at an appropriate temperature.
Levinthal's convincing arguments\cite{Levinthal1969-hi} (aka Levinthal's paradox) about proteins apply to RNA molecules as well, perhaps even more compellingly: arrival to thermal equilibrium should take a great deal more time than the biological lifetime of the organism. Evidently, mechanisms have evolved that contribute to a directed and efficient folding process, and quite possibly the active configuration is not at all like a sample from an equilibrium distribution, but rather a metastable state, provided only that it will typically last long enough for its purpose. 
Various measures of the difficulty in reaching a stable configuration, whether or not it is a true equilibrium,
are proposed based on the number of possible Watson-Crick or G$\cdot$U wobble pairings at each location. We call these measures ``ambiguity indexes,'' and focus on two, each of which is a function of both the primary and any given proposed secondary structure. Comparing families of RNA molecules that appear in ribonucleoproteins (``bound'' molecules) to RNA molecules that do not operate as nucleoproteins (``unbound'' molecules), we find that the unbound molecules have systematically lower ambiguity indexes, {\em provided that the indexes are computed using the comparative-analysis \cite{Gutell1992-hu} structure} (also known as the ``gold standard''). The effect largely disappears when the indexes are based, instead, on the so-called minimum-free energy (MFE) structures \cite{Mathews1999-hc,Zuker1999-rc}. Since MFE structures are derived from an approximation of the equilibrium distribution, we interpret these and other related results as evidence for metastable but not necessarily thermal equilibrium secondary structure.

%
%
%% Please keep the Author Summary between 150 and 200 words
%% Use first person. PLOS ONE authors please skip this step. 
%% Author Summary not valid for PLOS ONE submissions.

@@

% Use "Eq" instead of "Equation" for equation citations.
\section*{Author Summary}
Recent discoveries indicate that, in addition to being a messenger between DNA and protein, RNA molecules assume a wide range of biological functions. For biological macromolecules, structure is function. Experimental determination of RNA structures is still time-consuming, and computational approaches are of great importance. The prevailing computational approach tries to find the structure with the minimum energy, yet the relevance of this minimum energy structure to the native structure is still hotly debated. In this paper, we adopt a kinetic perspective, and argue that more emphasis should be placed on the folding process when trying to develop computational methods for RNA structure prediction. We present some statistical analyses using the primary and secondary data (sequence and base-pairs data) of RNA molecules, based on the concept of ``local ambiguity", i.e. the molecule's tendency to ``make a mistake" at a certain location when forming secondary structures. Our results highlight deficiencies in the minimum-energy approach, and demonstrate the importance of considering kinetics as well as protein-RNA interactions in developing computational approaches for RNA secondary structure prediction.

\section*{Introduction}

@@

Because of the time-consuming nature of experimental determination of RNA structures, a considerable amount of work has been put into computational (as opposed to experimental) approaches. For secondary structure prediction, comparative analysis is the gold standard  \cite{Gutell1992-hu}. But accurate comparative analysis typically requires a large number of homologous sequences, which may or may not be available. On the other hand, the prevailing {\em computational} approaches are based on an important and controversial assumption\cite{Flamm2008-ir}. Namely, that the structures of  non-coding RNA molecules, {\em in vivo}, are in thermal equilibrium. If this were the case, and if the configuration energies could be accurately specified, then secondary (and in principle, even tertiary) structures could be explored through Monte Carlo sampling from the Boltzmann equilibrium distribution. Alternatively, depending on the nature of the energy landscape, secondary structures could be approximated by identifying the minimum free energy (MFE) configuration  \cite{Mathews1999-hc,Zuker1999-rc}. However, even if we put aside the necessity for approximations, the biological relevance of equilibrium configurations has been a source of misgivings at least since 1969, when Levinthal pointed out that the time required to equilibrate might be too long by many orders of magnitude  \cite{Levinthal1969-hi}. In light of these observations, and considering the ``frustrated'' nature of the folding landscape, many have argued that when it comes to structure prediction for macromolecules, kinetic accessibility is more relevant than equilibrium thermodynamics  \cite{Higgs2000-gj,Flamm2008-ir,Baker1994-px}.

@@

We will introduce quantitative measures of ambiguity and demonstrate their statistical relationships to native secondary structure and to qualitative distinctions between RNA families.
Using a somewhat arbitrary convention, we will
refer to sequences of length four (four consecutive nucleotides) as {\em segments}. Keeping this convention in mind, each line of statistical
evidence is based on the same formal definition of the ``local ambiguity'' of a given segment. This is simply the number of its complementary pairs in the molecule. 
When we refer to the {\em location} of a segment, we will mean the location of the first element of the segment, counting from 5$^\prime$ to 3$^\prime$, and  
when we refer to the local ambiguity of a location, we will mean the local ambiguity of the segment at that location. 
Local ambiguity is an intrinsic property of the primary structure. We will be interested in exposing its statistical relationships to any given candidate secondary structure of the same molecule. For this purpose, for any particular secondary structure we distinguish three different kinds of locations:
\begin{description}
	\item[\textit{Single}:] Locations where all nucleotides in the corresponding segment are unpaired in the secondary structure;
	\item[\textit{Double}:] Locations where all nucleotides in the corresponding segment are paired in the secondary structure;
	\item[\textit{Transitional}:] Locations where some nucleotides in the corresponding segment are paired and others are unpaired in the secondary structure.
\end{description}
{\em Double} and {\em transitional} locations participate in the candidate secondary structure, while {\em single} locations do not.
As a general rule, local ambiguities measured at any of these locations will increase with the length of the molecule---there are simply more opportunities to find complementary sequences.
Instead of using the local ambiguities themselves, we focus on the {\em differences} between ambiguities in and around stems ({\em double} and {\em transitional} locations) from those at unpaired ({\em single}) locations. In particular, we defined the ``T-S ambiguity index'' to be the difference between the average local ambiguity at {\em transitional} locations minus the average at {\em single} locations. We defined the ``D-S ambiguity index'' similarly, but adjusted for an evident bias at {\em double} locations, most of which will have at least one complementary pair somewhere in the molecule. 
(The exceptions being bulges, which account for about 20\% of the {\em double} locations in our data set.)
With this in mind, we defined the D-S ambiguity index to be the average local ambiguity at those {\em double} locations that have at least one ambiguity minus the average over those {\em single} locations that have at least one ambiguity\footnote{
An argument could be made for including bulges in the definition of the {\em local ambiguity} in the first place. We have not explored this or other variations, though there is not shortage of arguably useful descriptive statistics for capturing the notion of ambiguity.}.

@@

\subsection*{A First Look at the Data: Shuffling Nucleotides}
Our goals are to explore connections between ambiguities and basic characteristics of RNA families, as well as the changes in these relationships, if any, when using comparative, as opposed to MFE, secondary structures.
For each molecule and each location $i$, the segment at $i$ has been assigned a ``local ambiguity'' $a_i(p)$ that depends only on the primary structure, and a label ({\it single}, {\it double}, or {\it transitional}) that depends only on the secondary structure. Since the local ambiguity, by itself, is strongly dependent on the length of the molecule, and possibly on other intrinsic properties,
we propose two {\em relative} ambiguity indexes: T-S and D-S, each of which depends on both the primary ($p$) and purported secondary ($s$) structures. Formally,
\begin{equation}
	d_\text{T-S}(p, s) = \frac
	{\sum_{j = 0}^{N - 3} a_j (p) 
	c^{\text{tran}}_j (s)}
	{\sum_{j = 0}^{N - 3} 
	c^{\text{tran}}_j (s)} - 
	\frac
	{\sum_{j = 0}^{N - 3} a_j (p) c_j^{\text{single}} (s)}
	{\sum_{j = 0}^{N - 3} c_j^{\text{single}} (s)}
\end{equation}
where we have used
$c_i^{\text{tran}}$ and $c_i^{\text{single}}$ for indicating whether location $i$ is {\it transitional} or {\it single} respectively. In other words, for each $i=1,2,\ldots,N-3$
\begin{align}
	c_i^{\text{tran}}(p, s) &=
	\begin{cases}
		1, & \text{if location $i$ is \textit{transitional}} \\
		0, & \text{otherwise}
	\end{cases} \\
	c_i^{\text{single}}(p, s) &=
	\begin{cases}
		1, & \text{if location $i$ is \textit{single}} \\
		0, & \text{otherwise}
	\end{cases}
\end{align}
In short, the T-S ambiguity index is the difference in the averages of the local ambiguities at {\em transitional} sites and {\em single} sites.
Since, as noted earlier, the local ambiguity at every {\em double} location, $i$, 
is likely to be at least one ($a_i\geq 1$), the D-S index involves only those {\em double} and {\em single} locations, $j$, for which $a_j\geq 1$:
\begin{equation}
	d_\text{D-S}(p, s) = \frac
	{\sum_{j = 0}^{N - 3} a_j (p) 
	\hat{c}^{\text{double}}_j (s)}
	{\sum_{j = 0}^{N - 3} 
	\hat{c}^{\text{double}}_j (s)} - 
	\frac
	{\sum_{j = 0}^{N - 3} a_j (p) \hat{c}_j^{\text{single}} (p,s)}
	{\sum_{j = 0}^{N - 3} \hat{c}_j^{\text{single}} (p,s)}
\end{equation}
Here, $\hat{c}_i^{\text{double}}$ and $\hat{c}^{\text{single}}_i$ indicates those {\em double} and {\em single} locations with at least one pair: for each $i=1,2,\ldots,N-3$,
\begin{align}
	\hat{c}_i^{\text{double}}(p, s) & =
	\begin{cases}
		1, & \text{if location $i$ is \textit{double} and $a_i(p)\geq 1$} \\
		0, & \text{otherwise}
	\end{cases} \\
	\hat{c}_i^{\text{single}}(p, s) & =
	\begin{cases}
		1, & \text{if location $i$ is \textit{single} and $a_i(p)\geq 1$} \\
		0, & \text{otherwise}
	\end{cases}
\end{align}

@@

To make this precise, consider an RNA molecule with primary structure $p$ and comparative secondary structure $s$. Construct a segment ``histogram function,'' $\mathcal{H}(p)$, which is the 
vector of the number of times that each of the $4^4$ possible segments appears in $p$.\footnote{${\cal H}$ is made a vector by fixing an arbitrary ordering of all possible segments.}.
Let $\mathcal{P}(p)$ be the set of all permutations of the ordering of nucleotides in $p$, and let 
$\mathcal{E}(p)\subseteq\mathcal{P}(p)$ be the subset of permutations that preserve the frequencies of four-tuples:
\[
{\cal E}(p) = \{p'\in\mathcal{P}(p): {\cal H}(p')={\cal H}(p)\}
\]
Clever algorithms exist for efficiently drawing independent samples from the uniform distribution on $\mathcal{E}$---see \cite{Kandel1996-kz, Fitch1983-ec, Altschul1985-nq}. Let $p^{(1)},
\ldots,p^{(K)}$ be $K$ such samples, and let $\displaystyle
d_\text{T-S}\big(p^{(1)},s\big),\ldots,d_\text{T-S}\big(p^{(K)},s\big)$ and
$\displaystyle
d_\text{D-S}\big(p^{(1)},s\big),\ldots,d_\text{D-S}\big(p^{(K)},s\big)$ 
be the corresponding T-S and D-S ambiguity indexes. The results are not very sensitive to $K$, nor to the particular sample, provided that $K$ is large enough. We used $K=$10,000. Finally, let 
$\alpha_\text{T-S}(p,s) \in [0,1]$ and
$\alpha_\text{D-S}(p,s) \in [0,1]$
be the left-tail empirical probabilities, under the distributions defined by the ensembles
\begin{align}
&  d_\text{T-S}(p,s),d_\text{T-S}\big(p^{(1)},s\big),\ldots,d_\text{T-S}\big(p^{(K)},s\big) 
\label{eqn:T-Sensemble}\\
\text{and \hglue .8cm} 
& d_\text{D-S}(p,s),d_\text{D-S}\big(p^{(1)},s\big),\ldots,d_\text{D-S}\big(p^{(K)},s\big) 
\label{eqn:D-Sensemble}
\end{align}
of choosing an ambiguity index less than or equal to $d_\text{T-S}(p,s)$ and 
$d_\text{T-S}(p,s)$, respectively:
\begin{align}
\alpha_\text{T-S}(p,s)  & = \frac
{1 + \#\{k\in\{1,\ldots,K\}: d_\text{T-S}\big(p^{(k)},s\big) \leq d_\text{T-S}(p,s)\}}{1+K} \label{eqn:alpha_T-S} \\
\alpha_\text{D-S}(p,s)  & = \frac
{1 + \#\{k\in\{1,\ldots,K\}: d_\text{D-S}\big(p^{(k)},s\big)\leq d_\text{D-S}(p,s)\}}{1+K} \label{eqn:alpha_D-S}
\end{align}
In essence, each $\alpha$ score is a self-calibrated ambiguity index.

@@

We are given a population of $M$ molecules, $m=1,\ldots,M$, each with a binary outcome measure $B_m\in\{-1,+1\}$.
There are two subpopulations of interest: the first $M_1$ molecules make up population 1 and the next $M_2$ molecules make up population 2; $M_1+M_2=M$. We observe $n_1$ plus values in population 1 and $n_2$ in population 2 
\begin{align}
n_1 & = \#\big\{m\in\{1,2,\ldots,M_1\}:B_m=+1\big\} \\
n_2 & = \#\big\{m\in\{M_1+1,M_1+2,\ldots,M\}:B_m=+1\big\}
\end{align}
We suspect that population 1 has less than its share of plus ones, meaning that the $n_1+n_2$ population of plus ones was not randomly distributed among the $M$ molecules. To be precise, let $N$ be the number of plus ones that appear from a draw, without replacement, of $M_1$ samples from $B_1,\ldots,B_M$. Under the null hypothesis, $H_o$, $n_1$ is a sample from the hypergeometric distribution on $N$:
\begin{equation}
\label{eqn:null}
\mathbb{P}\{N=n\} = \frac{\binom{M_1}{n}\binom{M_2}{n_1+n_2-n}}
{\binom{M}{n_1+n_2}}\quad \max\{0,n_1+n_2-M_2\}\leq n 
\leq \min\{n_1+n_2,M_1\}
\end{equation}
The alternative hypothesis, $H_a$, is that $n_1$ is too small to be consistent with $H_o$, leading to a left-tail test with p-value $\mathbb{P}\{N\leq n_1\}$ (which can be computed directly or using a statistical package, e.g. {\em hypergeom.cdf} in {\em scipy.stats}).

@@

The comparative secondary structure of an RNA molecule, when combined with its primary structure, can be used to construct a measure---the T-S ambiguity index---which distinguishes unbound from bound molecules with good accuracy. Can the same can be said for the D-S index?
Yes, albeit with lower accuracy. To demonstrate, we followed the identical procedure, except that we assigned the index $d_\text{D-S}(p,s)$ rather than $d_\text{T-S}(p,s)$ to each molecule.  The ROC curve (with area 0.65) is shown in the lower-left panel of Figure \ref{fig:UnboundVSBound}. The hypergeometric test, based on the signs of the D-S indexes
(positive counts, for each RNA family, can be found in Table \ref{table:NumberPositive}) has p-value $7.3\cdot 10^{-8}$. Finally, the right-hand panels in Figure \ref{fig:UnboundVSBound} mirror the left-hand panels, except that all secondary structures were computed by (approximately) minimizing free energy rather than comparative analysis. The classification results are substantially less convincing and the p-values substantially higher.

@@

Interestingly, there is a sharp difference in predicting the source of secondary structure in unbound molecules as opposed to bound molecules. Consider the top two ROC curves in Figure
\ref{fig:CompVSMFE}. In each of the two experiments a classifier was constructed by thresholding the T-S ambiguity index, declaring the secondary structure to be 
``comparative" when $d_\text{T-S}(p,s)<t$ and ``MFE" otherwise, since, as we have seen, smaller values of the ambiguity indexes are generally associated with the comparative secondary structure. The ROC curves are then swept out by varying $t$. The difference between the two panels is in the population used for the classification experiments---unbound molecules in the upper-left and bound molecules in the upper-right. In unbound molecules, small values of the T-S ambiguity index are a good indication that a secondary structure was derived from comparative rather than free-energy analysis, but not in bound molecules. The corresponding hypothesis tests seek evidence against the null hypotheses that in a given group (unbound or bound) the set of positive T-S ambiguity indexes ($d_\text{T-S}(p,s)>0$) are equally distributed between the comparative and free-energy derived indexes, and in favor of the alternatives that the T-S ambiguity indexes are less typically positive for the comparative secondary structures. The necessary data can be found in Table \ref{table:NumberPositive}. The results are consistent with the classification experiments: the hypergeometric p-value is 
is $5.4\cdot 10^{-14}$ for the unbound population and $0.07$ for the bound population.

@@

\section*{Discussion}
We have introduced two indexes designed as coarse measures of the prevalence of kinetic traps in the energy landscape of a structural RNA molecule. These indexes, which we call the T-S and D-S ambiguity indexes, are each a function of both the primary and any presumed secondary structure of the molecule. The first, the T-S index, compares the availability of false pairings (i.e. stems that are not part of the purported secondary structure) of sequences straddling transitions from double to single-stranded regions to the availability of false pairings strictly within single-stranded regions.
The second, the D-S index, results from a similar calculation, but using double-stranded sequences in place of {\em transitional} sequences. We have given evidence that either index can be used effectively to classify unbound RNA (molecules that do not act as components of ribonucleoproteins) from bound RNA (those that are components of ribonucleoproteins), {\em provided that the secondary structure is derived from comparative analysis}. 
Specifically, unbound molecules typically have lower ambiguity indexes than bound molecules. On the other hand, there is little evidence for this effect in the ambiguity indexes derived instead from MFE analyses, and classifiers based on MFE secondary structure are substantially less accurate. Furthermore, either index can be used in unbound molecules to effectively classify a given secondary structure as derived from comparative as opposed to MFE analysis, but not in bound molecules. In particular, among unbound molecules comparative-analysis structures typically have lower ambiguity indexes than MFE structures.
